\Large Last Writer Slices and Communication Traps

### 2013-10-31 paper.tex
paper changes

When programmers reason incorrectly, they permit some interleavings that should
be forbidden.  When such interleavings occur, program executions may fail,
experience performance degradation, or suffer other undesirable behavior.
Preventing such undesirable behavior is a fundamental challenge faced by
software engineers.  One important way to understand program behavior is to
monitor which read and write operations {\em depend} on one another.  Memory
operations are dependent if their order determines the values they read or
write.  Dependences are especially useful for understanding how threads
interact.

@@

We note that monitoring dependences is not only a fundamental part of the value
of such systems, but also a fundamental part of the {\em cost}.  Tracking
dependences requires executing additional monitoring code while the program
runs and many of the above systems have reported overheads too high for use in
production (10-100x
slowdown)~\cite{raceslicing,dmtracker,velodrome,recon,defuse,oshajava,chimera,coredet,stm}.
Unfortunately, many of these systems are most effective when they run in
production.

We argue in this work that dependence-tracking analysis should -- and can --
have low enough overhead for use in deployed systems.  The key to reducing the
overhead of dependence tracking is to carefully select and monitor only a
{\em subset} of dependences.  We start from two simple observations: (1) Reads are
frequent so we must minimize tracking overheads on reads, and (2) When
addressing concurrency, {\em inter-thread} dependences are most important.
Based on these observations, we focus on dependences that reflect {\em
communication} between threads.  Communication occurs when a thread reads or
overwrites a memory location that was most recently written by a different
thread; {\em i.e.}, inter-thread read-after-write (RAW) and write-after-write
(WAW).  Tracking only communication dependences preserves enough information to
implement useful applications, but has low enough overhead for always-on use in
production.  Framed by this idea, there are two parts to our work.

First, we describe {\em last writer slices}
(Section~\ref{sec:lastwriterslices}).  Last writer slices record the thread and
program point that last wrote each memory location.  Last writer slices are
useful because they expose the {\em provenance} of a memory location's contents
during debugging (Section~\ref{sec:debugging}).  When a failure occurs, last
writer slices are saved with a core dump.  Last writer slices are also
available during an execution in the debugger.  Tracking last writer slices has
low overhead because it only imposes overhead on write operations:  reads are
overhead-free.  In Section~\ref{sec:eval} we experimentally show
that collecting last writer slices has low overhead.

Second, on top of last writer slices we build {\em Communication Traps} or {\em
\ctraps}  (Section~\ref{sec:ctraps}).  \ctraps monitors inter-thread
communication with overhead low enough for production use in many important
cases.  When threads communicate, \ctraps delivers a {\em trap} that can be
handled by a trap handler.  Trap handlers implement {\em \ctraps applications}
through an extensible interface that can analyze or interpose on the execution
(Section~\ref{sec:apps}).  By default, \ctraps has very low overhead.  \ctraps
applications can perform arbitrary dependence analyses, but at their own
discretion and cost.  As we show in our evaluation, the overhead of \ctraps
applications scales with the complexity of their analysis.

@@

%Our system support for \ctraps incorporates the support for collecting last
%writer slices described in Section~\ref{sec:lastwriterslices}.  In addition,
%our \ctraps system support also monitors all read and write operations that
%might communicate and traps on all those that communicate.  
In addition to our base \ctraps design, we also investigate \ctrapsmm, a
variant of \ctraps that uses a flow-sensitive optimization to eliminate some
sources of overhead at a cost in analysis precision. \ctrapsmm tracks all write
operations that might communicate, but leaves some redundant read operations
untracked (\ctraps, with {\bf N}o {\bf R}edundant {\bf R}eads).

@@

\subsubsection{\ctrapsmm: Eliminating Redundant Instrumentation}
\label{sec:dpo}

The runtime support for \ctrapsmm is mostly the same as the runtime support for
our base \ctraps design, but \ctrapsmm adds trap hooks on only a subset of
reads.  \ctrapsmm's compiler support uses the program's control-flow structure
to identify reads that are likely to be {\em redundant}, and eliminates trap
hooks on those accesses.

We use dominator-based analysis to eliminate instrumentation that is likely to
be redundant.  Our analysis is function-local and works on code in SSA form.
The compiler looks at each pair of reads that are instrumented with trap hooks.
If a pair of instrumented reads both refer to a memory address stored in the
same SSA variable and the instrumentation points are {\em control equivalent},
then they are redundant.  Two reads, $A$ and $B$ are control equivalent if $A$
dominates $B$ and $B$ post-dominates $A$ -- every path to $B$ passes through
$A$, and every path passing through $A$ reaches $B$.  The latter read ("$B$")
is not instrumented.  Using this technique to remove instrumentation sites
preserves instrumentation along conditional control flow paths.  For example,
if the second of a pair of reads is conditionally executed, and the first is
not, then the pair are not control equivalent and both remain.

Figure~\ref{fig:insthooks} illustrates how a pair of control equivalent reads
are instrumented by the compiler for \ctrapsmm.  Pictured on the left is the
original program.  The shapes represent memory accesses with writes marked 'W'
and reads marked 'R'.  Notice that the the pentagon and the diamond are control
equivalent, despite the control flow structure separating them.  Pictured on
the right is the program after instrumentation.  There are two important things
to notice.  First, only the first of the control equivalent accesses are
instrumented, eliminating instrumentation overhead.  Second, the conditionally
executed accesses on either side of the branch are both instrumented.
Preserving these accesses is important for enabling flow-sensitive applications
using \ctrapsmm.

\begin{figure}[htb]
\centering
\includegraphics[width=.9\columnwidth]{figs/CTMMHooks.pdf}
\caption{\label{fig:insthooks}Placement of trap hooks for \ctrapsmm.}
\end{figure}

@@

\paragraph{\ctrapsmm}

\ctrapsmm has the same soundness guarantees as our base \ctraps design, but
compromises on completeness.  The concession in completeness is the result of
removing instrumentation on redundant reads.

\ctraps and \ctrapsmm rely on the same \lwt implementation, so all traps
delivered in \ctrapsmm are also delivered in the base \ctraps design.  Hence,
\ctrapsmm has the same soundness guarantee as basic \ctraps.  \ctrapsmm removes
trap hooks on some reads that have trap hooks in our base \ctraps design.
Without trap hooks, \ctrapsmm cannot deliver a trap before these reads, so it
is incomplete.  We designed our redundancy analysis to sacrifice completeness,
but preserve enough trap hooks to remain useful.  In \ctrapsmm, at least one of
each pair of control-equivalent accesses is instrumented with a trap hook.
Preserving at least one instrumenation site per control-equivalent pair
guarantees some monitoring still occurs (although the set of trapping accesses
may decrease substantially compared to our base \ctraps design).  Furthermore,
as Section~\ref{sec:dpo} discusses, instrumentation sites on interesting
control flow paths are preserved under our redundancy analysis.  For many
debugging applications, preserving this control flow information is important.

@@

We implemented our last writer slicing design, our base \ctraps designs, and
our \ctrapsmm design.  We built our compiler support as a plugin for GCC ({\tt
gcc-4.7.0 20111112}).  We used GCC's points-to and escape analysis support to
prune instrumentation to non-escaping memory locations.  We used GCC's
dominator and post-dominator computations to implement our redundancy analysis.
Our compiler instruments calls to {\tt free} and {\tt delete} as writes to the
pointer being deallocated.  Our analysis handles all code compilable using this
version of GCC except for a few cases: we do not handle accesses that compile
to GCC {\tt BITFIELD\_REFERENCE} IR types because these are undocumented, we do
not handle inlined assembly instructions, and we do not handle C++ exceptions.
Note that these limitations of our research prototype are not fundamental and
the limited cases are uncommon.

@@

The main performance result of our work is that our designs impose performance
overheads that are low enough for use in deployed systems.
Figure~\ref{fig:perfall} shows the slowdown suffered by our benchmarks due to last writer slicing
and \ctraps running with ``no-op'' trap hooks ({\em i.e.}, no
handlers), relative to the natively executing baseline.  These data reflect
our main performance findings.  We defer considering \ctrapsmm to
Section~\ref{sec:appperf} because no-op trap hooks have negligible overhead, so
redundancy analysis provides negligible improvement.  In applications that
perform analysis in read instrumentation, the benefit of redundancy analysis is
non-negligible.

@@

\subsubsection{Precision and Performance with \ctrapsmm}
We also evaluated the tradeoff of precision for performance of \ctrapsmm by
applying our redundancy analysis to these applications.  We compute the loss in
precision for CCI-Prev as the symmetric difference of the sets of points
reported by \ctraps and \ctrapsmm.  We compute the loss in precision for
communication graph collection as the symmetric difference in the edge lists
reported by \ctraps and \ctrapsmm.  Table~\ref{tab:precisionapps} shows our
precision results.

\begin{table}
\small
\centering
\begin{tabular}{c | r | r  }
\multirow{2}{*}{\bf Application} & \multicolumn{2}{c}{\bf \% Symmetric Difference} \\
& {\bf \em CCI-Prev} & {\bf \em Comm. Graph} \\ \hline
Apache& 0.2\% & 0.6\% \\
MySQL & 12.9\%  & 37.9\% \\
memcached  & 6.2\% & 4.3\% \\
LevelDB & 2.2\% & 9.9\% \\ \hline
blackscholes & 6.7\% & 0.0\% \\
dedup & 1.5\% & 3.5\% \\
canneal & 37.0\%  & 3.9\% \\
x264 & 0.0\% & 0.0\% \\
vips & 47.2\% & 11.3\% \\
ferret & 25.7\% & 11.2\%\\
fluidanimate & 1.4\% & 1.6\%\\
swaptions & 0.0\% & 0.5\%\\
streamcluster & 2.2\% & 3.8\%\\
\end{tabular}
\caption{\label{tab:precisionapps}The difference in precision for \ctraps applications using our base \ctraps design versus \ctrapsmm}
\end{table}

The results show that in many cases \ctrapsmm's redundancy analysis preserves
analysis information.  For CCI-Prev, {\tt swaptions} and {\tt x264} had
identical outputs under \ctrapsmm and our base \ctraps design.  The difference for other
benchmarks was often tolerable -- $0.2\%$ for Apache, $1.5$\% for {\tt dedup},
$1.4\%$ for {\tt fluidanimate}, and 2.2\% for LevelDB.  Results were similar
for communication graph collection.   The maximum difference in precision we saw (37.9\%)
was less the maximum for CCI-Prev (47.2\%).  Several applications saw
miniscule differences in their output ({\tt swaptions}, {\tt Apache}, {\tt
fluidanimate}) or had identical outputs under both schemes ({\tt blackscholes},
{\tt x264}).

We also quantified the performance improvement of \ctrapsmm.
Figure~\ref{fig:perfcci} shows the speedup of each application with \ctrapsmm
versus our base \ctraps design, as a percentage.  The speedups were low on average, with a
geometric mean speedup of around 4\% for communication graph collection.  The
key result here is that some applications that reaped larger performance
benefits ({\em e.g.} $>$10\% for {\tt x264}) had no precision loss (cf.
Table~\ref{tab:precisionapps}).  Others, like LevelDB had a modest performance
boost (about 7\%) at a modest precision cost (about 2\% for CCI-Prev, 10\% for graph collection).

These data show that if performance is paramount and imprecision is tolerable,
\ctrapsmm can provide marginally better overheads with little cost.  We note
that for applications with very high overhead, a 5-10\% improvement in
overheads is likely negligible.  However, for applications with very low
overhead to begin with ({\em e.g.} servers), a 5-10\%
performance improvement is more likely to matter.

\begin{figure}
\centering
\includegraphics[width=.9\columnwidth]{plots/appmm.pdf}
\caption{\label{fig:perfcci}Performance overheads of CCI-Prev and Graph Collection implemented with \ctrapsmm.}
\end{figure}

@@

\begin{table}
\renewcommand{\tabcolsep}{2pt}
\centering
\small
\begin{tabular}{l | r r | r r | r r}
\multirow{3}{*}{\bf Application} & \multicolumn{6}{c}{\bf Pruned Instrumentation} \\
               & \multicolumn{2}{c|}{\bf \em Static Local}          & \multicolumn{4}{c}{\bf \em  Redundant Reads}                 \\ 
               & \multicolumn{2}{c|}{\bf \em Operations}          & \multicolumn{2}{c|}{ \em  Static} & \multicolumn{2}{c}{ \em  Dynamic}                \\ \hline
Apache         &  21.4\%&(12K)                &  2.6\%& (1.4K)                        & 3.0\%   & (15.2M)                   \\
MySQL          &  21.3\%&(53K)                &  6.8\%& (17K)                         & 0.1\%   & (230M)                    \\
memcached      &  65.8\%&(7.5K)               & 1.4\% & (160)                        & $<$0.1\% & (10.2M)                   \\
LevelDB        &  44.0\%&(4.5K)               & 5.8\% & (598)                         & 2.2\%   & (6.8M)                    \\ \hline
blackscholes   &   7.5\%&(4)                  & 3.8\% & (2)                           & $<$0.1\%& (2)                       \\
dedup          &   22.0\%&(171)               & 0.1\% & (1)                           &  0.1\%  & (899K)                    \\
canneal        &   28.6\%&(226)               & 15.7\%& (124)                         &  15.2\% & (977M)                    \\
x264           &   29.8\%&(5960)              & 14.9\%& (2987)                        &  23.3\% & (869M)                    \\
vips           &   24.6\%&(9116)              & 5.2\% & (1919)                        &  16.0\% & (3.3B)                    \\
ferret         &   20.9\%&(792)               & 1.1\% & (40)                          &   1.6\% & (923M)                    \\
fluidanimate   &   19.6\%&(149)               & 0.4\% & (3)                           &   0.1\% & (138M)                    \\
streamcluster  &   11.6\%&(47)                & 5.4\% & (22)                          & $<$0.1\%& (5.8M)                    \\
swaptions      &   8.3\% &(19)                & 1.0\% & (2)                           & $<$0.1\%& (2)                       \\
\end{tabular}
\caption{\label{tab:char}Instrumentation pruned by static analysis.}
\end{table}

>>>>

When programmers reason incorrectly, they permit some interleavings that should
be forbidden.  When such interleavings occur, program executions may fail,
experience performance degradation, or suffer other undesirable behavior.
Preventing such undesirable behavior is a fundamental challenge faced by
software engineers.  An important way to understand program behavior is to
monitor which read and write operations depend on one another.  Memory
operations are dependent if their order determines the values they read or
write.  Inter-thread dependence analysis is especially useful for understanding
how threads interact.

@@

Many of the systems mentioned above are especially valuable when they run in
production systems on deployed code.  Unfortunately, tracking dependences is
not only a fundamental part of the value of these systems, but also a
fundamental part of their {\em cost}.  Tracking dependences requires executing
additional monitoring code while the program runs and many of the above systems
have reported overheads too high for use in production (10-100x
slowdown)~\cite{raceslicing,dmtracker,velodrome,recon,defuse,oshajava,chimera,coredet,stm}.

The key to reducing the overhead of dependence tracking is to carefully select
and monitor only a {\em subset} of dependences.  We start from two simple
observations: (1) Reads are frequent so we must minimize tracking overheads on
reads, and (2) In concurrency analyses, {\em inter-thread} dependences are most
important.  Based on these observations, we focus on dependences that reflect
{\em communication} between threads.  Communication occurs when a thread reads
or overwrites a memory location that was most recently written by a different
thread; {\em i.e.}, inter-thread read-after-write (RAW) and write-after-write
(WAW).  Tracking only communication dependences preserves enough information to
implement useful applications and analyses, but has low enough overhead for
always-on use in production systems.

Framed by these observations, we demonstrate in this work that an important,
versatile form of inter-thread dependence-tracking analysis has inherent
overheads that are low enough for use in deployed systems.  There are two parts
to our work:

First, we describe {\em last writer slices}
(Section~\ref{sec:lastwriterslices}).  Last writer slices record the thread and
program point that last wrote each memory location.     Last writer slices are
useful because they expose the {\em provenance} of a memory location's contents
during debugging (Section~\ref{sec:debugging}).  When a failure occurs, last
writer slices are saved with a core dump.  Last writer slices are also
available during an execution in the debugger.  Note that while prior work
({\em e.g.}, ~\cite{recon,fasttrack,conmem,conseq,cci}) has relied on {\em ad
hoc} last-writer analyses, in this work, we isolate the essential last-writer
information in the form of last writer slices and provide an implementation
with overheads appropriate for production systems.  Tracking last writer slices
has low overhead because it imposes overhead on only write operations:  reads
are overhead-free.  In Section~\ref{sec:eval} we experimentally show that
collecting last writer slices indeed has overheads suitable for production
systems for a variety of applications.

Second, on top of last writer slices we build {\em Communication Traps} or {\em
\ctraps}  (Section~\ref{sec:ctraps}).  \ctraps monitors inter-thread
communication with overhead low enough for production use in many important
cases.  When threads communicate, \ctraps delivers a software {\em trap} that can be
handled by a trap handler.  Trap handlers implement {\em \ctraps applications}
through an extensible interface that can analyze or interpose on the execution
(Section~\ref{sec:apps}).  By default, \ctraps has very low overhead.  \ctraps
applications can perform arbitrary dependence analyses, but at their own
discretion and cost.  As we show in our evaluation, the overhead of \ctraps
applications scales with the complexity of their analysis.

@@

%Our system support for \ctraps incorporates the support for collecting last
%writer slices described in Section~\ref{sec:lastwriterslices}.  In addition,
%our \ctraps system support also monitors all read and write operations that
%might communicate and traps on all those that communicate.

@@

@@

@@

We implemented last writer slicing and \ctraps.  We built our compiler support
as a plugin for GCC ({\tt gcc-4.7.0 20111112}).  We used GCC's points-to and
escape analysis support to prune instrumentation to non-escaping memory
locations.  We used GCC's dominator and post-dominator computations to
implement our redundancy analysis.  Our compiler instruments calls to {\tt
free} and {\tt delete} as writes to the pointer being deallocated.  Our
analysis handles all code compilable using this version of GCC except for a few
cases: we do not handle accesses that compile to GCC {\tt BITFIELD\_REFERENCE}
IR types because these are undocumented, we do not handle inlined assembly
instructions, and we do not handle C++ exceptions.  Note that these limitations
of our research prototype are not fundamental and the limited cases are
uncommon.

@@

The main performance result of our work is that our designs impose performance
overheads that are low enough for use in deployed systems.
Figure~\ref{fig:perfall} shows the slowdown suffered by our benchmarks due to last writer slicing
and \ctraps running with ``no-op'' trap hooks ({\em i.e.}, no
handlers), relative to the natively executing baseline.  These data reflect
our main performance findings.

@@

@@

\begin{table}
%\renewcommand{\tabcolsep}{2pt}
\centering
\small
\begin{tabular}{l | r r }
\multirow{3}{*}{\bf Application} & \multicolumn{2}{c}{\bf \em Pruned } \\
               & \multicolumn{2}{c}{\bf \em Static Local}  \\ 
               & \multicolumn{2}{c}{\bf \em Operations}    \\ \hline

Apache         &  21.4\%&(12K)                    \\
MySQL          &  21.3\%&(53K)                    \\
memcached      &  65.8\%&(7.5K)                   \\
LevelDB        &  44.0\%&(4.5K)                   \\ \hline
blackscholes   &   7.5\%&(4)                      \\
dedup          &   22.0\%&(171)                   \\
canneal        &   28.6\%&(226)                   \\
x264           &   29.8\%&(5960)                  \\
vips           &   24.6\%&(9116)                  \\
ferret         &   20.9\%&(792)                   \\
fluidanimate   &   19.6\%&(149)                   \\
streamcluster  &   11.6\%&(47)                    \\
swaptions      &   8.3\% &(19)                    \\
\end{tabular}
\caption{\label{tab:char}Instrumentation pruned by static analysis.}
\end{table}
